---
title: "Tools, Resources and Help Ideas"
subtitle: "draft"
format: html
# use quarto markdown to 
---


This page is to include pontential ideas to do for resources, tools, and ideas for time series



# .yml code to fix

1. _quarto.yml code. the sidebar code is different from time series and mine
    * time series sidebar does not appear on the side while my does. theme looks different
    * I think it just has to do w/ the theme
    
2.   back-to-top-navigation: true 
    * This is my original code and it works normally.
    * Time series uses this one: page-navigation: true 
      * When I use the time series I started getting this warning message
        * [WARNING] Could not fetch resource ../projects/project2.html
        * What is the difference between the two?
        

# using Definitons or footnotes

 The following is working code to use footnotes but works great for definitions. Obviously not the best way to reference def, but this seems like a simple and straight forward process. Seems super static so it wont take long to load as well.

White noise[^1]

[^1]: White noise consists of independent, identically distributed variables with mean 0, finite constant variance, and no correlation, representing purely random fluctuations over time.


# Streamline code

chapter_3_overview_code_draft.qmd is streamline

- Edits to variables happen in first code chunk and qmd runs all code for that chapter.

- May need to data frame set according to data


# color definitions 

so far I'm using the color purple for like definitions or the main topic of that paragraph but mostly definitions or explaining of a variable. 


I should make like a color code for the different colors I use.

- purple: purple is assign for variables or important subjects that are been define and or main topic of that paragraph and or just need to stand out. 

Pink: Pink will be assign to words that I don't know the definition and or meaning, and or words that need a definition link to them. This will help as I know that I can add the definition link later if I don't do it then




# add a concerns/doubs?

thinking of having a place where I can have my doubts and concerns, but maybe I can just do github issues for this since once it gets solve it will no longer be needed. doing issues will have more urgency to solve the problem.
ok, I'm going to do issues





# notes for my statistics journal


**Sampling Distributions**

Key points about modality and skewness in sampling distributions:

Modality:

- Unimodal: A sampling distribution with only one prominent peak.

- Bimodal: A sampling distribution with two distinct peaks.

- Multimodal: A sampling distribution with more than two peaks.

- Uniform: In a uniform distribution, every value within a specified range has an equal probability of occurring, creating a flat horizontal line when visualized on a graph.

Skewness:

- Positive Skewness (Right Skewed): The tail of the distribution extends further to the right side, meaning there are a few very high values compared to the majority of lower values.

- Negative Skewness (Left Skewed): The tail of the distribution extends further to the left side, indicating a few very low values compared to most higher values.

- Zero Skewness (Symmetrical): The distribution is balanced with the left and right sides mirroring each other. 



```{r}

library(ggplot2)

# Unimodal
ggplot(data.frame(x = rnorm(1000)), aes(x)) +
  geom_density(fill = "blue", alpha = 0.5) +
  ggtitle("Unimodal Distribution")

# Bimodal
ggplot(data.frame(x = c(rnorm(500, -2), rnorm(500, 2))), aes(x)) +
  geom_density(fill = "green", alpha = 0.5) +
  ggtitle("Bimodal Distribution")

# Multimodal
ggplot(data.frame(x = c(rnorm(300, -4), rnorm(300, 0), rnorm(300, 4))), aes(x)) +
  geom_density(fill = "purple", alpha = 0.5) +
  ggtitle("Multimodal Distribution")

# Uniform
ggplot(data.frame(x = runif(1000, -5, 5)), aes(x)) +
  geom_histogram(aes(y = ..density..), bins = 30, fill = "orange", color = "black", alpha = 0.5) +
  ggtitle("Uniform Distribution")


```





```{r}

# Positive Skew
ggplot(data.frame(x = rexp(1000, rate = 1)), aes(x)) +
  geom_density(fill = "red", alpha = 0.5) +
  ggtitle("Positive Skew (Right Skewed)")

# Negative Skew
ggplot(data.frame(x = -rexp(1000, rate = 1)), aes(x)) +
  geom_density(fill = "cyan", alpha = 0.5) +
  ggtitle("Negative Skew (Left Skewed)")

# Symmetrical
ggplot(data.frame(x = rnorm(1000)), aes(x)) +
  geom_density(fill = "blue", alpha = 0.5) +
  ggtitle("Zero Skewness (Symmetrical)")


```



Key Differences Between General and Sampling Distributions
----------------------------------------------------------

| **Aspect**            | **General Distributions**                                               | **Sampling Distributions**                                                     |
|------------------------|-------------------------------------------------------------------------|--------------------------------------------------------------------------------|
| **Focus**             | Distribution of raw data or random variables                           | Distribution of a statistic (e.g., sample mean, proportion)                   |
| **Population or Sample** | Describes population or a single sample                               | Derived from repeated sampling of a population                                 |
| **Examples**          | Normal, Uniform, Exponential                                           | Sampling distribution of sample mean, proportion, etc.                        |
| **Derived From**      | Directly observed data                                                 | Repeatedly calculated from samples                                            |
| **Shape**             | Depends on the data (e.g., normal, skewed)                             | Depends on sample size and population (Central Limit Theorem often applies)   |
| **Role in Statistics**| Describe the data or population characteristics                        | Used for inferential statistics (e.g., confidence intervals, hypothesis testing) |


