---
title: "Chapter 3 r code examples and practice"
subtitle: ""
format: 
  html:
    error: false
    message: false
    warning: false
    embed-resources: true
    toc: true
    code-fold: false
---


```{r, echo=FALSE, message=FALSE, warning=FALSE}

# Loading R packages
if (!require("pacman")) install.packages("pacman")
pacman::p_load(tidyverse,
               tsibble, fable,
               feasts, tsibbledata,
               fable.prophet,
               patchwork,
               lubridate,
               rio,
               ggplot2,
               kableExtra
               )
```

This qmd is made to summarize chapter 3 and have models for the chapter to view.

-   I have a goal to make the code not be reliant on adjusting variable names and titles.

    -   This will require having an initial r chunk that assigns variable names and dataset.

    -   I will still have multiple df.

-   This will be a bit hard and the models will have to be pretty general because some data sets required different approaches

    -   I think maybe I can also assign if its yearmonth, yearquater etc, but that will require so many if statements.

        -   Idk how I'm going to approach this

```{r}
# Assign your column names to variables
df0 <- rio::import("https://byuistats.github.io/timeseries/data/constructionequip_manu_orders_shipments.csv")


dates <- "date"
x_col <- "constructionequip_ord"
y_col <- "constructionequip_ship"



# Assign plot labels
x_label <- "Month"
y_label <- "New Orders & Value of Equip"
plot_title <- "Time Series of Construction Equip: New Orders & Equipment"
```

```{r}


df <- df0 |>
  mutate(
    date = lubridate::mdy(.data[[dates]]),
    x = as.numeric(.data[[x_col]]), # Convert and rename to x
    y = as.numeric(.data[[y_col]])  # Convert and rename to y
  ) |>
  select(date, x, y) 


```

```{r, error=FALSE, message=FALSE, warning=FALSE}

df1 <- df |> # this makes a new df so either df before or this one is use. 
  mutate(obs = row_number()) |> # makes new column with periods
  select(date, obs, x, y)

dfx <- df |> # lone df for variable x = ord
  mutate(obs = row_number()) |> 
  select(date, x)

dfy <- df |> # lone df for y = ship
  mutate(obs = row_number()) |> 
  select(date, y)
```

test

```{r, error=FALSE, message=FALSE, warning=FALSE}
#| label: time series plot of both variables
# this is code for hw 3-1

# this code is same as the dfx2 & dfy2 code
df1 <- df1 |>
  mutate(index = tsibble::yearmonth(date)) |> # 3.1
  as_tsibble(index = index) |>
  select(index, date, x, y)



autoplot(df1, .vars = x) +
  geom_line(data = df1, aes(x = index, y = y), color = "#E69F00") +
  labs(
    x = x_label,
    y = y_label,
    title = plot_title
  ) +
  theme(plot.title = element_text(hjust = 0.5))
```
