---
title: "Chapter overview and Task"
subtitle: "Chapter 1"
format: 
  html:
    error: false
    message: false
    warning: false
    embed-resources: true
    toc: true
    code-fold: true
---

This is the overview of the chapter 4. I will add info from the index file here. I will filter from there to what will go here.
My goal is to first have the website have a map of the concepts in time series. I want to first avoid doing it by chapters since it seperates topics and limit my learning to that lesson. Bigger picture first.





# code for lesson 1.5

the following does the multiplicative model

```{r}
source("../common_functions.R")


```



```{r}

# Set random seed for reproducibility
set.seed(123) 

# Set parameters & initialize vectors
num_years <- 10
n <- 12 * num_years
sigma <- .75
a <- 0.03
b <- 1
c <- 0.5 
trend <- seasonal <- x_t <- rep(0,n)
time_seq <- seq(1,n)

# Generate correlated error terms
w <- rnorm(n + 4, 0.2, 0.1) # Changed to a mean of 1 and sd of 0.03
z = w + lead(w,1) + lead(w,2) + lead(w,3) + lead(w,4)
z  = head(z, n)

# Get date
year_seq <- lubridate::year(today()) - num_years  + (time_seq - 1) %/% 12
month_seq <- (time_seq - 1) %% 12 + 1
date_seq <- ymd(paste0(year_seq,"-",month_seq,"-01"))

# Get data
for (t in 1:n) {
  trend[t] <- exp(a * t)
  seasonal[t] <- exp( b * sin(t / 12 * 2 * pi * 1)  + c * cos(t / 12 * 2 * pi * 3) + 1 )
  x_t[t] <- trend[t] * seasonal[t] * z[t] # Note R's definition of the mult. model
}

x_df <- data.frame(x_t = x_t, trend = trend, seasonal = seasonal)

start_year <- lubridate::year(today()) - num_years
start_date <- lubridate::ymd(paste0(start_year,"-01-01"))

# start_date <- lubridate::ymd("1958-01-01")
date_seq <- seq(start_date,
    start_date + months(nrow(x_df)-1),
    by = "1 months")

x_df_ts <- x_df |>
  mutate(
    date = date_seq,
    month = tsibble::yearmonth(date)
  ) |>
  select(date, month, trend, seasonal, x_t) |>
  as_tsibble(index = month)
```

```{r}
#| echo: false

x_df_ts |>
  rename(
    Date = date,
    Month = month,
    "Trend, $$m_t$$" = trend,
    "Seasonal, $$s_t$$" = seasonal,
    "Data, $$x_t$$" = x_t
  ) |>
  display_partial_table(6,2)
```






```{r}

trend_plot <- ggplot(x_df_ts, aes(x=month, y=trend)) + 
  geom_line() +
  labs(
    title="Plot of Trend", 
    x="Month", 
    y="Trend"
    ) +
  theme(plot.title = element_text(hjust = 0.5))

seasonal_plot <- ggplot(x_df_ts, aes(x=month, y=seasonal)) + 
  geom_line() +
  labs(
    title="Plot of Seasonal Effect", 
    x="Month", 
    y="Seasonal"
    ) +
  theme(plot.title = element_text(hjust = 0.5))

error_plot <- ggplot(x_df_ts, aes(x = month, y = x_t / trend / seasonal)) + 
  geom_line() +
  labs(
    title="Plot of Random Error Term", 
    x="Month", 
    y="Random"
  ) +
  theme(plot.title = element_text(hjust = 0.5))

x_plot <- ggplot(x_df_ts, aes(x=month, y=x_t)) + 
  geom_line() +
  labs(
    title="Plot of Simulated Time Series", 
    x="Month", 
    y="x_t"
  ) +
  theme(plot.title = element_text(hjust = 0.5))

x_plot <- x_plot  + labs(title = "True (Simulated) Values", x = NULL)
trend_plot <- trend_plot + labs(title = NULL, x = NULL)
seasonal_plot <- seasonal_plot + labs(title = NULL, x = NULL)
error_plot <- error_plot + labs(title = NULL)

x_plot / trend_plot / seasonal_plot / error_plot 
```










Random Error Term (Error Plot):
$$ e_t = \frac{x_t}{m_t \cdot s_t} $$

Trend:
$$ m_t = e^{a \cdot t} $$

Seasonal:
$$ s_t = e^{b \cdot \sin\left(\frac{2 \pi t}{12}\right) + c \cdot \cos\left(\frac{6 \pi t}{12}\right) + 1} $$

Combined Time Series (Multiplicative Model):
$$ x_t = m_t \cdot s_t \cdot z_t $$




